---
title: Unit Sizes in the American Civil War
---

```{r,results='hide',echo=FALSE,message=FALSE}
library("dplyr")
library("rjson")
```

For many battles and sources of data about them, the number of personnel present (or engaged) is either not given or not known. Instead, the units involved are given. Somtimes the CWSAC simply states that "brigades" or "corps" were involved. This program generates the number of personnel associated with each unit size. Unfortunately, there is cons  iderable uncertainty about the number of personnel in a unit, both due to organizational differences and attrition. Thus this program generates a probability distribution over the number of personnel in a unit.

The best source for this is Eicher (p. 65, 74) who Heitman (1903):

- Army: Composed of 1-8 corps, most frequently 3.
- Corps: Composed of 2-6 divisions, most frequently 3.
- Division: 

    - Infantry: 2-5 brigades, most frequently 3
    - Cavalry: 2-6 brigades, most frequently 4.
    
- Brigade: 

    - Infantry: 2-12 regiments, most frequently 4
    - Cavalry: 2-8 regiments, most frequently 4
    
- Regiment: This was the fundamental unit for mustering.

   - Their initial size varied from 845-1010, 866-1046, or 830-1612 men (src 52). This was a regimental staff + 10 companies.
   - Artillery regiment: average 1800 men
   - Heavy artillery regiments average 1200 men in 12 companies. These were often converted to infantry units instead of being assigned to garrison duty.
   - Cavalry regiments: 4-6 squadrons, each squadron was composed of 2 companies, for a total of 660-1168 men (src 304).
- Battalion: No official size, but "every part of a regiment, composed of two or more companies will be designated a battalion ... "

   - Regiments could have between 2 and 5 battaltions, but usually a battalion was half a regiment.
   - Cavalry equivalent was a squadron. 
   
- Company: 

    - Infantry: 64-83 privates and 19 officers for a total of 83-102 men.
    - Cavalry (troop): 79-105 men
    - Artillery (battery): 80-156 men (4-6 guns)
    
With attrition regiments were often at much lower than their official strength. They could be between 300-400 in mid war [wikipedia](http://en.wikipedia.org/wiki/Infantry_in_the_American_Civil_War).

In terms of attrition the Confederate and Union armies differed in an important respect. In the Union army recruits were assigned to new regiments and thus the regimental size would continue to decrease until it was no longer effective. In the Confederate army, recruits were assigned to existing regiments.

```{r}
units_table <- read.csv("eicher_units_table.csv",
                        check.names = TRUE, stringsAsFactors = FALSE)
```

Union infantry unit sizes:

```{r,results='asis',echo=FALSE}
knitr::kable(filter(units_table, 
       belligerent == "Union",
       as.logical(infantry))
       %>% select(-belligerent, -infantry)
    )
```

Union cavalry unit sizes
```{r,results='asis',echo=FALSE}
knitr::kable(filter(units_table, 
       belligerent == "Confederate",
       ! as.logical(infantry))
       %>% select(-belligerent, -infantry)
    )
```

Confederate infantry unit sizes:

```{r,results='asis',echo=FALSE}
knitr::kable(filter(units_table, 
       belligerent == "Confederate",
       as.logical(infantry))
       %>% select(-belligerent, -infantry)
    )
```

Confederate cavalry unit sizes
```{r,results='asis',echo=FALSE}
knitr::kable(filter(units_table, 
       belligerent == "Confederate",
       ! as.logical(infantry))
       %>% select(-belligerent, -infantry)
    )
```

Also see: 

- [Infantry in the American Civil War](http://en.wikipedia.org/wiki/Infantry_in_the_American_Civil_War) in en.wikipedia.org.
- [Cavalry in the American Civil War](http://en.wikipedia.org/wiki/Cavalry_in_the_American_Civil_War) in en.wikipedia.org.
- [Field Artillery in the American Civil War](http://en.wikipedia.org/wiki/Field_artillery_in_the_American_Civil_War)
- National Park Service, [From Regiment to President](http://www.nps.gov/resources/story.htm?id=299): The Structure and Command of Civil War Armies
- http://www.civilwarhome.com/armyorganization.htm
- Eicher, John H. and David J. Eicher, *Civil War High Commands*.
- Heitman, Francis B. (1903) *Historical Register and Dictionary of the United States Army* [Google books](http://books.google.com/books?id=6RLMX6OOgsAC)
    
Simulate distributions of personnel size for military units in the Civil War

Often  the number of units is known, but the exact number of troops is not given or known.
The following code simulates the size of those units using uncertainty about the number of subunits and the number of troops in a regiment (given attrition).

In usage, these may scaled by the "effective" size multipliers used in Livermore (1900), p. 69. The Union army did not consistently report "effective" forces, while the Confederates appeared to do a better job of reporting effective forces.

Strength of regiments (Livermore 1900, p. 67).
Regiment of 1050 includes 70 non-combatants (10 staff, 20 musicians, 40 band, wagoners, men detailed to HQ, quartermaster of medical).

Average regimental present for duty

- Shiloh: 560
- Fair Oaks: 650
- Chancellorsville: 530
- Gettysburg: 375
- Chickamagua: 440
- Wilderness: 440
- Sherman's army in May 1864: 305

Based on these values assume that regiments had between 30-65% of their statutory size.

```{r}
REGIMENT_SIZE <- 1050
PCT_UNIT_LOW <- 305 / 1050
PCT_UNIT_HIGH <- 650 / 1050
```


```{r }
#library("triangle")
set.seed(276184053)
N <- 1000
```

# Functions

The distributions will be summarized by the mean, sd, and 5th, 25th, 50th, 75th, and 95th percentiles.

```{r }
summarize_data <- function(x) {
    summarize(data.frame(x = x),
              mean = mean(x),
              sd = sd(x),
              p75 = quantile(x, 0.75),
              p25 = quantile(x, 0.25),
              median = median(x),
              p05 = quantile(x, 0.05),
              p95 = quantile(x, 0.95)
              )
}
```

Generate "summary" statistics if the exact size of a unit is known.

```{r }
summarize_data_exact <- function(n) {
    data.frame(mean = n,
               median = n,
               sd = 0,
               p05 = n, p95 = n,
               p25 = n, p25 = n)
}
```

If the the size of the unit is distributed uniform, then return the exact summary statistics of the distribution.

```{r }
uniform_size <- function(low, high) {
    range <- high - low
    data_frame(mean = 0.5 * (low + high),
               median = 0.5 * (low + high),
               sd = sqrt((1 / 12) * (low - high)^2),
               p05 = 0.05 * range + low,
               p25 = 0.25 * range + low,
               p75 = 0.75 * range + low,
               p95 = 0.95 * range + low)

}
```

Eicher summarizes the distribution of number of sub-units within a unit by the min, max, mean and mode. I'll draw from a binomial which matches the mean of the support given in that table.

```{r}
draw_number_subunits <- function(n, low, high, mean) {
  size <- high - low
  prob <- (mean - low) / size
  low + rbinom(n, size, prob)  
}

```

The size of a regiment is drawn from a uniform distribution between low and high.

```{r }
draw_unif <- function(n, low, high) {
    sample.int(high - low, n, replace = TRUE) + low
}
```

To calculate the size of a brigades and higher, draw the number of sub-units from a binomial distribution, and for each sub-unit draw the size.


```{r}
draw_unit_personnel <- function(n, low, high, mean, .FUN) {
    subunits <- draw_number_subunits(n, low, high, mean)
    vapply(subunits, function(n) sum(.FUN(n)), 0)
}

draw_unit_personnel_unif <- function(n, low, high, .FUN) {
    subunits <- low + sample.int(high - low, n, replace = TRUE)
    vapply(subunits, function(n) sum(.FUN(n)), 0)
}

draw_unit_personnel_exact <- function(n, k, .FUN) {
    vapply(rep(n, k), function(n) sum(.FUN(n)), 0)
}
```

The list `unit_sizes` will store the output.
```{r}
unit_sizes <- list(union = list(), confed = list())
```

# Union Army

## Infantry

Union companies were 100 men, but assume the field unit size percent from Livermore, so companies are more like 30-65 men.

```{r }
UNION_COMPANY_SIZE <- 100
unit_sizes[["union"]][["infantry_company"]] <-
  uniform_size(PCT_UNIT_LOW * UNION_COMPANY_SIZE,
               PCT_UNIT_HIGH * UNION_COMPANY_SIZE)
              
```

For the size of regiments, use the range of values given by Livermore.

```{r }
UNION_REGIMENT_SIZE <- 1050

union_regiment_low <- PCT_UNIT_LOW * UNION_REGIMENT_SIZE
union_regiment_high <- PCT_UNIT_HIGH * UNION_REGIMENT_SIZE
```
The distribution is uniform and the summary stats can be calculated exactly.
```{r}
unit_sizes[["union"]][["infantry_regiment"]] <-
  uniform_size(union_regiment_low, union_regiment_high)
```
However, I'll need to draw random values from it for other unit sizes.
```{r}
draw_union_infantry_regiments <- function(n) draw_unif(n, union_regiment_low, union_regiment_high)
```

Draw brigades, divisions, corps and armies using the values from the tables of sub-units:

```{r}
draw_union_infantry_brigades <- function(n) {
  sz <- filter(units_table, belligerent == "Union", as.logical(infantry), unit.type == "regiments per brigade")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_infantry_regiments)                    
}
unit_sizes[["union"]][["infantry_brigade"]] <-
  summarize_data(draw_union_infantry_brigades(N))

draw_union_infantry_divisions <- function(n) {
  sz <- filter(units_table, belligerent == "Union", as.logical(infantry), unit.type == "brigades per division")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_infantry_brigades)
}
unit_sizes[["union"]][["infantry_division"]] <-
  summarize_data(draw_union_infantry_divisions(N))

draw_union_infantry_corps <- function(n) {
  sz <- filter(units_table, belligerent == "Union", as.logical(infantry), unit.type == "divisions per corps")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_infantry_divisions)                    
}
unit_sizes[["union"]][["infantry_corps"]] <-
  summarize_data(draw_union_infantry_corps(N))

draw_union_armies <- function(n) {
  sz <- filter(units_table, belligerent == "Union", as.logical(infantry), unit.type == "divisions per corps")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_infantry_corps)                    
}
unit_sizes[["union"]][["army"]] <-
  summarize_data(draw_union_armies(N))
```

## Cavalry

Assume a cavalry company is the same size as an infantry company
```{r}
unit_sizes[["union"]][["cavalry_company"]] <-
  unit_sizes[["union"]][["cavalry_company"]]

draw_union_cavalry_companies <-
  function(n) draw_unif(n, PCT_UNIT_LOW * 100, PCT_UNIT_HIGH * 100)
```

A cavalry squadrom is 2 companies
```{r}
draw_union_cavalry_squadrons <-
  function(n) {
     draw_unit_personnel_exact(n, 2, draw_union_cavalry_companies)
  }
unit_sizes[["union"]][["cavalry_company"]] <-
  summarize_data(draw_union_cavalry_squadrons(N))
```

Assume cavalry regiment is the same size as an infantry regiment
```{r}
draw_union_cavalry_regiments <- draw_union_infantry_regiments

unit_sizes[["union"]][["cavalry_regiment"]] <-
  unit_sizes[["union"]][["infantry_regiment"]]
```

Draw brigades, divisions, corps and armies using the values from the tables of sub-units:

```{r}
draw_union_cavalry_brigades <- function(n) {
  sz <- filter(units_table, belligerent == "Union", !as.logical(infantry),
               unit.type == "regiments per brigade")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_cavalry_regiments)                    
}
unit_sizes[["union"]][["cavalry_brigade"]] <-
  summarize_data(draw_union_cavalry_brigades(N))

draw_union_cavalry_divisions <- function(n) {
  sz <- filter(units_table, belligerent == "Union", !as.logical(infantry),
               unit.type == "brigades per division")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_cavalry_brigades)                    
}
unit_sizes[["union"]][["cavalry_division"]] <-
  summarize_data(draw_union_cavalry_divisions(N))

draw_union_cavalry_corps <- function(n) {
  sz <- filter(units_table, belligerent == "Union", !as.logical(infantry),
               unit.type == "divisions per corps")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_union_cavalry_divisions)                    
}
unit_sizes[["union"]][["cavalry_corps"]] <-
  summarize_data(draw_union_cavalry_corps(N))

```

## Artillery

# Confederate Unit Sizes

## Infantry

Confederate companies were 100 men, but assume the field unit size percent from 
Livermore, so companies are more like 30-65 men.

```{r }
CONFED_COMPANY_SIZE <- 100
unit_sizes[["confederate"]][["infantry_company"]] <-
  uniform_size(PCT_UNIT_LOW * UNION_COMPANY_SIZE,
               PCT_UNIT_HIGH * UNION_COMPANY_SIZE)
              
```

For the size of regiments, use the range of values given by Livermore.

```{r }
CONFED_REGIMENT_SIZE <- 1050

confed_regiment_low <- PCT_UNIT_LOW * CONFED_REGIMENT_SIZE
confed_regiment_high <- PCT_UNIT_HIGH * CONFED_REGIMENT_SIZE
```
The distribution is uniform and the summary stats can be calculated exactly.
```{r}
unit_sizes[["confed"]][["infantry_regiment"]] <-
  uniform_size(confed_regiment_low, confed_regiment_high)
```
However, I'll need to draw random values from it for other unit sizes.
```{r}
draw_confed_infantry_regiments <-
  function(n) draw_unif(n, confed_regiment_low, confed_regiment_high)
```

Draw brigades, divisions, corps and armies using the values from the tables of sub-units:

```{r}
draw_confed_infantry_brigades <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", as.logical(infantry),
               unit.type == "regiments per brigade")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_infantry_regiments)                    
}
unit_sizes[["confed"]][["infantry_brigade"]] <-
  summarize_data(draw_confed_infantry_brigades(N))

draw_confed_infantry_divisions <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", as.logical(infantry),
               unit.type == "brigades per division")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_infantry_brigades)
}
unit_sizes[["confed"]][["infantry_division"]] <-
  summarize_data(draw_confed_infantry_divisions(N))

draw_confed_infantry_corps <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", as.logical(infantry),
               unit.type == "divisions per corps")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_infantry_divisions)                    
}
unit_sizes[["confed"]][["infantry_corps"]] <-
  summarize_data(draw_confed_infantry_corps(N))

draw_confed_armies <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", as.logical(infantry),
               unit.type == "divisions per corps")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_infantry_corps)                    
}
unit_sizes[["confed"]][["army"]] <-
  summarize_data(draw_confed_armies(N))
```

## Cavalry

Assume a cavalry company is the same size as an infantry company
```{r}
unit_sizes[["confed"]][["cavalry_company"]] <-
  unit_sizes[["confed"]][["cavalry_company"]]

draw_confed_cavalry_companies <-
  function(n) draw_unif(n, PCT_UNIT_LOW * 100, PCT_UNIT_HIGH * 100)
```

A cavalry squadrom is 2 companies
```{r}
draw_confed_cavalry_squadrons <-
  function(n) {
     draw_unit_personnel_exact(n, 2, draw_confed_cavalry_companies)
  }
unit_sizes[["confed"]][["cavalry_company"]] <-
  summarize_data(draw_confed_cavalry_squadrons(N))
```

Assume cavalry regiment is the same size as an infantry regiment
```{r}
draw_confed_cavalry_regiments <- draw_confed_infantry_regiments

unit_sizes[["confed"]][["cavalry_regiment"]] <-
  unit_sizes[["confed"]][["infantry_regiment"]]
```

Draw brigades, divisions, corps and armies using the values from the tables of
sub-units:

```{r}
draw_confed_cavalry_brigades <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", !as.logical(infantry), 
               unit.type == "regiments per brigade")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_cavalry_regiments)                    
}
unit_sizes[["confed"]][["cavalry_brigade"]] <-
  summarize_data(draw_confed_cavalry_brigades(N))

draw_confed_cavalry_divisions <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", !as.logical(infantry), 
               unit.type == "brigades per division")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_cavalry_brigades)                    
}
unit_sizes[["confed"]][["cavalry_division"]] <-
  summarize_data(draw_confed_cavalry_divisions(N))

draw_confed_cavalry_corps <- function(n) {
  sz <- filter(units_table, belligerent == "Confederate", !as.logical(infantry), 
               unit.type == "divisions per corps")
  draw_unit_personnel(n, sz$low, sz$high, sz$avg, draw_confed_cavalry_divisions)                    
}
unit_sizes[["confed"]][["cavalry_corps"]] <-
  summarize_data(draw_confed_cavalry_corps(N))

```


## Artillery

# Data

What do we need? 

- infantry: company, regiment, brigade, division, corps, army
- cavalry: company, regiment, brigade, divions, corps
- artillery: battery, section / company, brigade, regiment



```{r }
con <- file("unit_sizes.json", "w")
cat(toJSON(unit_sizes),
    file = con)
close(con)
```

# TODO

- Should companies be at less than full strength? When regiments had casualties,
  did each regiment lose personnel, or were the personnel rearranged into fewer
  companies.
