# Unit Sizes in the American Civil War

For many battles and sources of data about them, the number of personnel present (or engaged) is either not given or not known. Instead, the units involved are given. Somtimes the CWSAC simply states that "brigades" or "corps" were involved. This program generates the number of personnel associated with each unit size. Unfortunately, there is considerable uncertainty about the number of personnel in a unit, both due to organizational differences and attrition. Thus this program generates a probability distribution over the number of personnel in a unit.

The best source for this is Eicher (p. 65, 74)

- Army: Composed of 1-8 corps, most frequently 3.
- Corps: Composed of 2-6 divisions, most frequently 3.
- Division: 

    - Infantry: 2-5 brigades, most frequently 3
    - Cavalry: 2-6 brigades, most frequently 4.
    
- Brigade: 

    - Infantry: 2-12 regiments, most frequently 4
    - Cavalry: 2-8 regiments, most frequently 4
    
- Regiment: This was the fundamental unit for mustering.

   - Their initial size varied from 845-1010, 866-1046, or 830-1612 men (src 52). This was a regimental staff + 10 companies.
   - Artillery regiment: average 1800 men
   - Heavy artillery regiments average 1200 men in 12 companies. These were often converted to infantry units instead of being assigned to garrison duty.
   - Cavalry regiments: 4-6 squadrons, each squadron was composed of 2 companies, for a total of 660-1168 men (src 304).
- Battalion: No official size, but "every part of a regiment, composed of two or more companies will be designated a battalion ... "

   - Regiments could have between 2 and 5 battaltions, but usually a battalion was half a regiment.
   - Cavalry equivalent was a squadron.
   
   
- Company: 

    - Infantry: 64-83 privates and 19 officers for a total of 83-102 men.
    - Cavalry (troop): 79-105 men
    - Artillery (battery): 80-156 men (4-6 guns)
    
With attrition regiments were often at much lower than their official strength. They could be between 300-400 in mid war [wikipedia](http://en.wikipedia.org/wiki/Infantry_in_the_American_Civil_War).


In terms of attrition the Confederate and Union armies differed in an important respect. In the Union army recruits were assigned to new regiments and thus the regimental size would continue to decrease until it was no longer effective. In the Confederate army, recruits were assigned to existing regiments.

Also see: 

- [Infantry in the American Civil War](http://en.wikipedia.org/wiki/Infantry_in_the_American_Civil_War) in en.wikipedia.org.
- [Cavalry in the American Civil War](http://en.wikipedia.org/wiki/Cavalry_in_the_American_Civil_War) in en.wikipedia.org.
- [Field Artillery in the American Civil War](http://en.wikipedia.org/wiki/Field_artillery_in_the_American_Civil_War)
- National Park Service, [From Regiment to President](http://www.nps.gov/resources/story.htm?id=299): The Structure and Command of Civil War Armies
- http://www.civilwarhome.com/armyorganization.htm
- Eicher, John H. and David J. Eicher, *Civil War High Commands*.
    
   
# Regiment

Confederate infantry field unit components (Eicher, p. 73)

======================= === ==== ==== ======  
unit type               low high avg  mode
----------------------- --- ---- ---- ------
corps per army          1   4    2.74 2
divisions per corps     2   7    3.10 3
brigades per division   2   7    3.62 4
regiments per brigade   2   20   4.71 5
======================= === ==== ==== ======

Confederate cavalry field unit components (Eicher, p. 73)

======================= === ==== ==== ======  
unit type               low high avg  mode
----------------------- --- ---- ---- ------
corps per army          2   3    2.43 2
divisions per corps     2   7    2.81 2
brigades per division   2   8    4.21 4
======================= === ==== ==== ======

Union infantry field unit components (Eicher, p. 66)

======================= === ==== ==== ======  
unit type               low high avg  mode
----------------------- --- ---- ---- ------
corps per army          1   8    3.71 3
divisions per corps     2   6    2.91 3
brigades per division   2   5    2.80 3
regiments per brigade   2   12   4.73 4
======================= === ==== ==== ======

Union cavalry field unit components (Eicher, p. 66)

======================= === ==== ==== ======  
unit type               low high avg  mode
----------------------- --- ---- ---- ------
divisions per corp      2   6    2.92 3
brigades per division   2   6    2.55 2
regiments per brigade   2   8    3.90 4
======================= === ==== ==== ======

Simulate distributions of personnel size for military units in the Civil War

Often  the number of units is known, but the exact number of troops is not given or known.
The following code simulates the size of those units using uncertainty about the number of subunits and the number of troops in a regiment (given attrition).

In usage, these may scaled by the "effective" size multipliers used in Livermore (1900), p. 69. The Union army did not consistently report "effective" forces, while the Confederates appeared to do a better job of reporting effective forces.

- Union artillery and infantry: 0.93
- Union cavalry: 0.85
- Proportion of Union officers: 4--7%
- Proportion of Confederate officers: 6.5--11%

Strength of regiments (Livermore 1900, p. 67).
Regiment of 1050 includes 70 non-combatants (10 staff, 20 musicians, 40 band, wagoners, men detailed to HQ, quartermaster of medical).

Average regimental present for duty

- Shiloh: 560
- Fair Oaks: 650
- Chancellorsville: 530
- Gettysburg: 375
- Chickamagua: 440
- Wilderness: 440
- Sherman's army in May 1864: 305

In the OR, numbers of "effectives" in infantry divisions or corps is 89--93 percent of number "present for duty", while in the cavalry these numbers are 83--86 percent.



```{r }
library("plyr")
library("jsonlite")
#library("triangle")
set.seed(276184053)
N <- 500
```

# General Functions

The distributions will be summarized by the mean, sd, and 5th, 25th, 50th, 75th, and 95th percentiles.

```{r }
summarize_data <- function(x) {
    summarize(x,
              mean = mean(troops),
              median = median(troops),
              sd = sd(troops),
              p75 = quantile(troops, 0.75),
              p25 = quantile(troops, 0.25),
              p05 = quantile(troops, 0.05),
              p95 = quantile(troops, 0.95)
              )
}
```

If the exact size of the unit is known (or assumed).

```{r }
exact_size <- function(n) {
    data.frame(mean = n,
               median = n,
               sd = 0,
               p05 = n, p95 = n,
               p25 = n, p25 = n)
}
```

If the the size of the unit is distributed uniform, then return the exact values:

```{r }
uniform_size <- function(low, high) {
    range <- high - low
    data.frame(mean = 0.5 * (low + high),
               median = 0.5 * (low + high),
               sd = 0,
               p05 = 0.05 * range + low,
               p25 = 0.25 * range + low,
               p75 = 0.75 * range + low,
               p95 = 0.95 * range + low)

}
```

Special functions to draw the standard units in the infantry.

The number of troops in a regiment is drawn from a uniform distribution
between low and high.

```{r }
draw_regiment <- function(n = 1, low = 300, high = 650) {
    data.frame(troops = round(runif(n, low, high)))
}

draw_brigade <- function(low = 2, high = 12, mode = 4, regiment = draw_regiment) {
    rng <- high - low
    n <- low + rbinom(1, rng, (mode - low) / (rng + 1))
    ## n <- round(rtriangle(1, a = low, b = high, c = mode), 0)
    troops <- sum(regiment(n))
    data.frame(regiments = n, troops = troops)
}

draw_division <- function(low = 2, high = 5, mode = 3, brigade = draw_brigade, ...) {
    #n <- round(rtriangle(1, a = low, b = high, c = mode), 0)
    rng <- high - low
    n <- low + rbinom(1, rng, (mode - low) / (rng + 1))
    ret <- colwise(sum)(rdply(n, brigade(...)))
    ret$.n <- NULL
    ret$brigades <- n
    ret
}

draw_corps <- function(low = 2, high = 6, mode = 3, division = draw_division, ...) {
    rng <- high - low
    n <- low + rbinom(1, rng, (mode - low) / (rng + 1))
    ## n <- round(rtriangle(1, a = low, b = high, c = mode), 0)
    ret <- colwise(sum)(rdply(n, division(...)))
    ret$.n <- NULL
    ret$divisions <- n
    ret
}

draw_army <- function(low = 1, high = 8, mode = 3, corps = draw_corps, ...) {
    rng <- high - low
    n <- low + rbinom(1, rng, (mode - low) / (rng + 1))
    ## n <- round(rtriangle(1, a = low, b = high, c = mode), 0)
    ret <- colwise(sum)(rdply(n, corps(...)))
    ret$.n <- NULL
    ret$corps <- n
    ret
}
```


Ships complements. Get distribution from the complements of all ships found in the
Union and Confederate navies on dbpedia.



```{r }
ships <- read.csv("ships.csv")
## Remove outliers and older ships.
ship_complements <-
    summarize(data.frame(troops = with(ships, complement[complement > 10 & complement < 500])),
          mean = mean(troops),
          median = median(troops),
          sd = sd(troops),
          p75 = quantile(troops, 0.75),
          p25 = quantile(troops, 0.25),
          p05 = quantile(troops, 0.05),
          p95 = quantile(troops, 0.95)
          )
```

# Union Army

## Infantry

Union companies were 100 men, but assume 30-65%.

```{r }
union_company <- uniform_size(30, 65)
```


Union regiments were 1000, but on the field due to attrition ranged between 300 and 400

```{r }
draw_union_regiment <- function(n) draw_regiment(n, 300, 650)
union_regiment <- uniform_size(300, 650)
```


Brigade: 2-12 regiments per brigade, mode = 4

```{r }
draw_union_brigade <- function() {
    draw_brigade(low = 2, high = 12, mode = 4,
                 regiment = draw_union_regiment)
}
union_brigade <- summarize_data(rdply(N, draw_union_brigade))
```


Division: 2-5 brigades per division, mode = 3

```{r }
draw_union_division <- function() {
    draw_division(low = 2, high = 5, mode = 3,
                  brigade = draw_union_brigade)
}
union_division <- summarize_data(rdply(N, draw_union_division))
```


Corps: 2-6 divisions per corps, mode = 3

```{r }
draw_union_corps <- function () {
    draw_corps(low = 2, high = 6, mode = 3,
               division = draw_union_division)
}
union_corps <- summarize_data(rdply(N, draw_union_corps))
```


Armies: 1-8 corps per army, mode = 3

```{r }
draw_union_army <- function() {
    draw_army(low = 1, high = 8, mode = 3,
              corps = draw_union_corps)

}
union_army <- summarize_data(rdply(N, draw_union_army))
```


## Cavalry

Union cavalry regiment is the same size as an infantry regiment.

```{r }
draw_union_cav_regiment <- function(n) draw_regiment(n, 300, 650)
union_cav_regiment <- uniform_size(300, 650)
```


Brigade: 2-20 regiments per brigade, mode = 5

```{r }
draw_union_cav_brigade <- function() {
    draw_brigade(low = 2, high = 4, mode = 3,
                 regiment = draw_union_cav_regiment)
}
union_cav_brigade <- summarize_data(rdply(N, draw_union_cav_brigade))
```


Division: 2-7 brigades per division, mode = 4

```{r }
draw_union_cav_division <- function() {
    draw_division(low = 2, high = 4, mode = 3,
                  brigade = draw_union_cav_brigade)
}
union_cav_division <- summarize_data(rdply(N, draw_union_cav_division))
```


Cavalry corps had 3 divisions.

```{r }
draw_union_cav_corps <- function() {
    draw_corps(low = 3, high = 3, mode = 3,
               brigade = draw_union_cav_division)
}
union_cav_corps <- summarize_data(rdply(N, draw_union_cav_corps))
```


## Artillery

Union artillery regiment is the same size as infantry regiment

```{r }
union_artillery_regiment <- uniform_size(300, 650)
```


Section = 2 guns, battery = 6 guns; brigades = 5 batteries

```{r }
union_artillery_section <- exact_size(2 * 12 + 1)
union_artillery_battery <- exact_size(6 * 12 + 1)
union_artillery_brigade <- exact_size((6 * 12 + 1) * 5 + 1)
```



## Ships


```{r }
union_ships <- ship_complements
```


# Confederate



## Infantry


Companies of 100, assume 30-65% fill

```{r }
confed_company <- uniform_size(30, 65)
```


Confed regiments were 1000, but on the field due to attrition ranged between 300 and 400

```{r }
draw_confed_regiment <- function(n) draw_regiment(n, 300, 400)
confed_regiment <- uniform_size(300, 650)
```


Brigade: 2-20 regiments per brigade, mode = 5

```{r }
draw_confed_brigade <- function() {
    draw_brigade(low = 2, high = 12, mode = 5,
                 regiment = draw_confed_regiment)
}
confed_brigade <- summarize_data(rdply(N, draw_confed_brigade))
```


Division: 2-7 brigades per division, mode = 4

```{r }
draw_confed_division <- function() {
    draw_division(low = 2, high = 7, mode = 4,
                  brigade = draw_confed_brigade)
}
confed_division <- summarize_data(rdply(N, draw_confed_division))
```


Corps: 2-6 divisions per corps, mode = 3

```{r }
draw_confed_corps <- function () {
    draw_corps(low = 2, high = 12, mode = 3,
               division = draw_confed_division)
}
confed_corps <- summarize_data(rdply(N, draw_confed_corps))
```


Armies: 1-4 corps per army, mode = 2

```{r }
draw_confed_army <- function() {
    draw_army(low = 1, high = 4, mode = 2,
              corps = draw_confed_corps)
}
confed_army <- summarize_data(rdply(N, draw_confed_army))
```


## Cavalry


Cavalry regiment, the same size as infantry regiment

```{r }
draw_confed_cav_regiment <- function(n) draw_regiment(n, 300, 650)
confed_cav_regiment <- uniform_size(300, 650)
```


Brigade: 2-20 regiments per brigade, mode = 5

```{r }
draw_confed_cav_brigade <- function() {
    draw_brigade(low = 2, high = 4, mode = 3,
                 regiment = draw_confed_cav_regiment)
}
confed_cav_brigade <- summarize_data(rdply(N, draw_confed_cav_brigade))
```


Division: 2-7 brigades per division, mode = 4

```{r }
draw_confed_cav_division <- function() {
    draw_division(low = 2, high = 4, mode = 3,
                  brigade = draw_confed_cav_brigade)
}
confed_cav_division <- summarize_data(rdply(N, draw_confed_cav_division))
```


Cavalry corps assumed same size as infantry

```{r }
draw_confed_cav_corps <- function() {
    draw_corps(low = 2, high = 12, mode = 3,
               brigade = draw_confed_cav_division)
}
confed_cav_corps <- summarize_data(rdply(N, draw_confed_cav_corps))
```


## Artillery

Union artillery regiment is the same size as infantry regiment

```{r }
confed_artillery_regiment <- uniform_size(300, 650)
```


Section = 2 guns, battery = 4 guns; brigades = 5 batteries

```{r }
confed_artillery_section <- exact_size(2 * 12 + 1)
confed_artillery_battery <- exact_size(4 * 12 + 1)
confed_artillery_brigade <- exact_size((4 * 12 + 1) * 5 + 1)
```


## Ships


```{r }
confed_ships <- ship_complements
```


# Data


```{r }
organization_sizes <-
    list(union =
         list(
             company = union_company,
             regiment = union_regiment,
             brigade = union_brigade,
             division = union_division,
             corps = union_corps,
             army = union_army,
             cavalry_regiment = union_cav_regiment,
             cavalry_brigade = union_cav_brigade,
             cavalry_division = union_cav_division,
             cavalry_corps = union_cav_corps,
             artillery_regiment = union_artillery_regiment,
             artillery_section = union_artillery_section,             
             artillery_battery = union_artillery_battery,
             artillery_brigade = union_artillery_brigade,
             ship = union_ships),
         confederate =
         list(
             company = confed_company,
             regiment = confed_regiment,
             brigade = confed_brigade,
             division = confed_division,
             corps = confed_corps,
             army = confed_army,
             cavalry_regiment = confed_cav_regiment,
             cavalry_brigade = confed_cav_brigade,
             cavalry_division = confed_cav_division,
             cavalry_corps = confed_cav_corps,
             artillery_regiment = confed_artillery_regiment,
             artillery_section = confed_artillery_section,             
             artillery_battery = confed_artillery_battery,
             artillery_brigade = confed_artillery_brigade,
             ship = confed_ships)
         )
             
con <- file("unit_sizes.json", "w")
cat(toJSON(organization_sizes),
    file = con)
close(con)
```
